---
title: "Simulation examples for MPQ distances"
output: tufte::tufte_html
toc: false
vignette: >
  %\VignetteIndexEntry{gentry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```


# Setup

```{r}
library(ape)
library(mpqDist)
library(plotly)
library(phyloseq)
library(tidyverse)
library(viridis)
library(phangorn)
library(patchwork)
data(gentry)
gentry
```

# Simulations

We're going to test whether our null means and variances look right.
To do this, we will generate "OTU tables" that are independent normal random variables, with potentially different standard deviations.
We'll use the gentry data as a skeleton (same tree, same number of samples and sites, same standard deviations for the species when we look into changing standard deviations).
```{r}
sim_homo = gentry
sim_hetero = gentry
n = nsamples(gentry)
p = ntaxa(gentry)
sds = apply(log(1 + otu_table(gentry)), 2, sd)
sim_homo_matrix = matrix(rnorm(n = n * p), nrow = n, ncol = p)
rownames(sim_homo_matrix) = sample_names(gentry)
colnames(sim_homo_matrix) = taxa_names(gentry)
sim_hetero_matrix = sweep(sim_homo_matrix, MARGIN = 2, STATS = sds, FUN = '*')
otu_table(sim_homo) = otu_table(sim_homo_matrix, taxa_are_rows = FALSE)
otu_table(sim_hetero) = otu_table(sim_hetero_matrix, taxa_are_rows = FALSE)
```

## Homoskedastic case

```{r}
base_sample = 1
nmv = get_null_mean_and_variance(sim_homo)
mpq_distances = get_mpq_distances(otu_table(sim_homo), phy_tree(sim_homo))
animation_df = make_animation_df(mpq_distances, get_avg_distances_to_set, means_and_vars = nmv, base_sample, sample_data(sim_homo))
p = ggplot(aes(x = Lat, y = avg_dist, color = Elev), data = animation_df) +
    geom_point(aes(frame = frame)) +
    geom_hline(aes(yintercept = null_mean, frame = frame)) +
    geom_hline(aes(yintercept = null_mean - 2 * null_sd, frame = frame)) +
    geom_hline(aes(yintercept = null_mean + 2 * null_sd, frame = frame)) +
    stat_smooth(aes(frame = frame), se = FALSE) +
    scale_x_reverse() +
    scale_color_viridis()
ggplotly(p)

p = ggplot(aes(x = Lat, y = avg_dist, color = Elev), data = animation_df) +
    geom_point(aes(frame = frame)) +
    geom_hline(aes(yintercept = median, frame = frame)) +
    geom_hline(aes(yintercept = upper, frame = frame)) +
    geom_hline(aes(yintercept = lower, frame = frame)) +
    stat_smooth(aes(frame = frame), se = FALSE) +
    scale_x_reverse() +
    scale_color_viridis()
ggplotly(p)
```


## Heteroskedastic case


```{r}
base_sample = 1
nmv = get_null_mean_and_variance(sim_hetero)
mpq_distances = get_mpq_distances(otu_table(sim_hetero), phy_tree(sim_hetero))
animation_df = make_animation_df(mpq_distances, get_avg_distances_to_set, means_and_vars = nmv, base_sample, sample_data(sim_hetero))
p = ggplot(aes(x = Lat, y = avg_dist, color = Elev), data = animation_df) +
    geom_point(aes(frame = frame)) +
    geom_hline(aes(yintercept = null_mean, frame = frame)) +
    geom_hline(aes(yintercept = null_mean - 2 * null_sd, frame = frame)) +
    geom_hline(aes(yintercept = null_mean + 2 * null_sd, frame = frame)) +
    stat_smooth(aes(frame = frame), se = FALSE) +
    scale_x_reverse() +
    scale_color_viridis()
ggplotly(p)


p = ggplot(aes(x = Lat, y = avg_dist, color = Elev), data = animation_df) +
    geom_point(aes(frame = frame)) +
    geom_hline(aes(yintercept = median, frame = frame)) +
    geom_hline(aes(yintercept = upper, frame = frame)) +
    geom_hline(aes(yintercept = lower, frame = frame)) +
    stat_smooth(aes(frame = frame), se = FALSE) +
    scale_x_reverse() +
    scale_color_viridis()
ggplotly(p)
```

## Non-null distributions


The two-gradient simulation described in the paper:
```{r}
make_anticorrelated_noise <- function(n, tr) {
    Q = vcv(tr)
    p = ncol(Q)
    Qeig = eigen(Q)
    D = Qeig$values
    D[1:(p/2)] = 0
    D = D / sum(D)
    D = diag(D)
    E = matrix(rnorm(n * p), nrow = n, ncol = p)
    E %*% sqrt(D) %*% t(Qeig$vectors)
}


set.seed(0)
p = 300
n = 100
tr = rtree(p)
alt = sample(c(-1, 1), n, replace = TRUE)
precip = sample(c(-1, 1), n, replace = TRUE)

phy_big = 1:p %in% Descendants(tr)[[469]]
phy_small_u = 1:p %in% Descendants(tr)[[503]]
phy_small_d = 1:p %in% Descendants(tr)[[529]]
M = .5 * outer(alt, phy_big) + outer(precip, phy_small_u) - outer(precip, phy_small_d)
X = M + rnorm(n = n * p, mean = 10, sd = .75) + p * .25 * make_anticorrelated_noise(n, tr)
colnames(X) = tr$tip.label
two_effect_sim = phyloseq(otu_table(X, taxa_are_rows = FALSE), phy_tree(tr), sample_data(data.frame(alt = alt, precip = precip)))
#nmv = get_null_mean_and_variance(two_effect_sim)
mpq_distances = get_mpq_distances(otu_table(two_effect_sim), phy_tree(two_effect_sim))
animation_df_alt = make_animation_df(mpq_distances, get_avg_distances_to_set, means_and_vars = nmv, base_sample = which.min(alt), sample_data(two_effect_sim))
animation_df_precip = make_animation_df(mpq_distances, get_avg_distances_to_set, means_and_vars = nmv, base_sample = which.min(alt), sample_data(two_effect_sim))
## p1 = ggplot(aes(x = factor(alt), y = avg_dist), data = animation_df_alt) +
##     geom_boxplot(aes(frame = frame))
## ggplotly(p1)

## p3 = ggplot(aes(x = factor(precip), y = avg_dist), data = animation_df_precip) +
##     geom_boxplot(aes(frame = frame))
## ggplotly(p3)
```

```{r}
math_labels <- c(
  "0" = "r = 0",
  "0.5" = "r = .5",
  "1" = "r = 1"
)
p1 = ggplot(subset(animation_df_alt, (frame %in% c(0,1) | (frame >= .49 & frame <= .51)) & avg_dist > 0)) +
    geom_boxplot(aes(x = factor(alt), y = avg_dist)) +
    facet_wrap(~ as.factor(frame), scales = "free_y", labeller = labeller(.default = math_labels)) +
    scale_x_discrete(breaks = c(-1, 1), labels = c("low", "high")) +
    xlab("Altitude")
p2 = ggplot(subset(animation_df_precip, (frame %in% c(0,1) | (frame >= .49 & frame <= .51)) & avg_dist > 0)) +
    geom_boxplot(aes(x = factor(precip), y = avg_dist)) +
    facet_wrap(~ as.factor(frame), scales = "free_y", labeller = labeller(.default = math_labels)) +
    scale_x_discrete(breaks = c(-1, 1), labels = c("wet", "dry")) +
    xlab("Moisture")
h_patch <- p1 / p2 & ylab(NULL) & theme(plot.margin = margin(5.5, 5.5, 5.5, 2))
#pdf("two-factor-sim.pdf", width = 6, height = 4)
wrap_elements(h_patch) +
  labs(tag = "MPQr distance to reference sample") +
  theme(
    plot.tag = element_text(size = rel(1), angle = 90),
    plot.tag.position = "left"
  )
#dev.off()
```


## More systematic, and in the tree basis
```{r}
convert_to_tree_basis <- function(mu, Qeig) {
    mutilde <- t(Qeig$vectors)%*% mu
}

compare_to_null_tree_basis <- function(mutilde, r, d_noise, Qeig) {
    dr = mpqDist:::get_dr(Qeig, r)
    w = dr * d_noise
    exp_val = t(mutilde * dr) %*% mutilde
    variance = 2 * sum(w^2)
    return(exp_val / sqrt(variance))
}

get_maximizing_r <- function(mutilde, d_noise, Qeig, rvec = mpqDist:::r_transform(0:100/100)) {
    ratios = sapply(rvec, function(r) compare_to_null_tree_basis(mutilde, r, d_noise, Qeig))
    return(rvec[which.max(ratios)])
}
```

```{r}
p = ntaxa(gentry)
Q = vcv(phy_tree(gentry))
Qeig = eigen(Q)
best_r_vec = numeric(p)
for(i in 1:p) {
    mutilde = rep(0, p)
    mutilde[i] = 1
    ##d_noise = c(rep(0, p/2), rep(1, p/2)) + .01
    ##d_noise = Qeig$values^(-2) / sum(Qeig$values^(-2))
    d_noise = rep(1, p)
    d_noise = d_noise / sum(d_noise)
    best_r_vec[i] = get_maximizing_r(mutilde, d_noise, Qeig)
}
ggplot(data.frame(best_r = best_r_vec, index = 1:p)) +
    geom_point(aes(x = index, y = best_r))

```


## Make an example with the perfect binary tree

```{r}
k = 8
p = 2^k
tr = ape::stree(n = 2^k, type = "balanced")
tr$edge.length = rep(1, nrow(tr$edge))
Q = ape::vcv(tr)
Qeig = eigen(Q)
clades = Descendants(tr)
best_r_vec = numeric(length(clades))
clade_sizes = numeric(length(clades))
for(i in 1:length(clades)) {
    mu = 1:p %in% clades[[i]]
    mu = mu / sqrt(sum(mu^2))
    mutilde = convert_to_tree_basis(mu, Qeig)
    ##d_noise = c(rep(0, p/2), rep(1, p/2)) + .01
    ##d_noise = Qeig$values^(-2) / sum(Qeig$values^(-2))
    d_noise = rep(1, p)
    d_noise = d_noise / sum(d_noise)
    best_r_vec[i] = get_maximizing_r(mutilde, d_noise, Qeig)
    clade_sizes[i] = length(clades[[i]])
}
p_perfect = ggplot(data.frame(best_r = best_r_vec, clade_fraction = clade_sizes / p)) +
    scale_x_log10() +
    xlab("Clade fraction") +
    ylab("Best r") +
    geom_point(aes(x = clade_fraction, y = best_r)) +
    ggtitle("Perfect binary tree")
```

## Same as above but with the Gentry tree


```{r}
tr = phy_tree(gentry)
p = ntaxa(gentry)
Q = ape::vcv(tr)
Qeig = eigen(Q)
clades = Descendants(tr)
best_r_vec = numeric(length(clades))
clade_sizes = numeric(length(clades))
time_since_divergence_of_mrca = numeric(length(clades))
tree_height = Q[1,1]
for(i in 1:length(clades)) {
    if(length(clades[[i]]) > 1) {
        mrca = getMRCA(tr, clades[[i]])
    } else {
        mrca = clades[[i]]
    }
    parent_of_mrca = tr$edge[which(tr$edge[,2] == mrca),1]

    mu = 1:p %in% clades[[i]]
    mu = mu / sqrt(sum(mu^2))
    mutilde = convert_to_tree_basis(mu, Qeig)
    ##d_noise = c(rep(0, p/2), rep(1, p/2)) + .01
    ##d_noise = Qeig$values^(-2) / sum(Qeig$values^(-2))
    d_noise = rep(1, p)
    d_noise = d_noise / sum(d_noise)
    best_r_vec[i] = get_maximizing_r(mutilde, d_noise, Qeig)
    clade_sizes[i] = length(clades[[i]])
    if(length(parent_of_mrca) > 0) {
        time_since_divergence_of_mrca[i] = tree_height - adephylo::distRoot(tr, parent_of_mrca)
    } else{
        time_since_divergence_of_mrca[i] = 0
    }
}
p_gentry =
    ggplot(data.frame(best_r = best_r_vec, clade_fraction = clade_sizes / p)) +
    geom_point(aes(x = clade_fraction, y = best_r)) +
    scale_x_log10() +
    xlab("Clade fraction") +
    ylab("Best r") +
    stat_smooth(aes(x = clade_fraction, y = best_r), method = "gam",  se = FALSE) +
    ggtitle("Gentry dataset tree")

size_one_df = subset(data.frame(best_r = best_r_vec, time_since_divergence_of_mrca), clade_sizes == 1)
size_one_df$Group = as.vector(tax_table(gentry)[,"group"])
p_size_one = ggplot(size_one_df) +
    geom_point(aes(x = time_since_divergence_of_mrca, y = best_r, shape = Group), na.rm = TRUE) +
    scale_shape(na.translate = FALSE) + ylab("Best r") + xlab("Distance to parent node") +
    ggtitle("Individual species in Gentry dataset tree")
```

```{r}
pdf("best-r-by-clade-size.pdf", width = 7, height=5)
(p_perfect + p_gentry) / p_size_one
dev.off()
```


```{r}
p = 2^k
otu_tab_perfect = matrix(rnorm(n * p, sd = 1 / sqrt(p)), nrow = n, ncol = p)
colnames(otu_tab_perfect) = tr$tip.label
## this is a clade of size 203, so roughly half of the tree, appears to be mainly Magnoliales
target_clade = 1:p %in% 1:(p/128)
target_clade = target_clade / sum(target_clade^2)
tree_additive_effect = outer(seq(-1,1,length.out=nrow(otu_tab_perfect)), target_clade)
otu_tab_perfect = otu_tab_perfect + tree_additive_effect
sim_perfect_tree = phyloseq(phy_tree(tr), otu_table(otu_tab_perfect, taxa_are_rows = FALSE),
                            sample_data(data.frame(tree_score = seq(-1, 1, length.out = n))))
base_sample = 1
nmv = get_null_mean_and_variance(sim_perfect_tree)
mpq_distances = get_mpq_distances(otu_table(sim_perfect_tree), phy_tree(sim_perfect_tree))
animation_df = make_animation_df(mpq_distances, get_avg_distances_to_set, means_and_vars = nmv, base_sample, sample_data(sim_perfect_tree))

p = ggplot(aes(x = tree_score, y = (avg_dist - null_mean) / null_sd), data = animation_df) +
    geom_point(aes(frame = frame)) +
#    geom_hline(aes(yintercept = null_mean, frame = frame)) +
#    geom_hline(aes(yintercept = null_mean - 2 * null_sd, frame = frame)) +
#    geom_hline(aes(yintercept = null_mean + 2 * null_sd, frame = frame)) +
    stat_smooth(aes(frame = frame), se = FALSE) +
    scale_x_reverse() +
    scale_color_viridis() 
ggplotly(p)
```
